---
title: "Identification of drought prone sites"
author: "William H MacKenzie"
date: "31/10/2024"
format:
  typst:
    toc: false
    toc-depth: 1
    toc-title: Contents
    section-numbering: 1.1.1
    columns: 1
editor: source
execute:
  echo: false
  error: false
  warning: false
  message: false
  fig.width: 6
  fig.height: 4
  fig.align: 'center'
  fig.cap: true
  fig.pos: H
  out.width: '100%'
  dev: pdf
  fig.ext: pdf
  cache: false
  fig.retina: 2
  dpi: 600
  fig.asp: 1.5
  fig.path: "./figures/"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(foreach)
require(ggplot2)
require (ggthemes)
require(tidyverse)
require(data.table)
library(flextable)
source("_3Dbarplot_Function.R")
```

## Introduction
 This document is intended to provide a summary of the analysis and results for the project "Identifying drought prone sites where plantation failure without best practices should be expected under current climate conditions". The analysis has two objectives the following objectives:

1.  Identifying drought prone sites where plantation failure without best practices should be expected under current climate conditions. Initial analysis uses the historic normal period (1961-1990) and compared to the modern climate ()

2.  Identifying site series that are likely to experience planting failure under drought events of different severity. For this report using the 2021 heat-dome year as an example catastrophic drought year.


```{r import prepped data, echo=FALSE, message = FALSE, warning=FALSE}
rsmr_cmd.long <- fread("./outputs/modelled_aSMR_by_rSMR_long.csv")
fordrat <- fread("./outputs/rSMR_aSMR_DroughtModel.csv")
fordrat.long <- fordrat  %>% pivot_longer(cols = -"BGC", names_to = "rSMR", values_to = "aSMR.fordrat") %>%  mutate(bgc = BGC) %>% select(bgc, rSMR, aSMR.fordrat)
fwrite(fordrat.long, file = "./outputs/fordrat_by_rSMR_long.csv")
klinka <- fread("./inputs/klinka_tree_asmr_mins_orig.csv") %>% pivot_longer(cols = -"spp", names_to = "frequency", values_to = "min_aSMR") 
asmr.codes <- fread("./inputs/asmr_codes.csv")

```

## Determining drought prone sites from site adjusted CMD
We use 4 sources of information to determine drought prone sites. The first two are the site series suitability table and the edatopic tables informing CCISS to find the driest rSMR for each tree species in each BGC. This provides a measure of the drought tolerance of each tree species in each BGC.
The third is a CMD modelled aSMR x rSMR grid. Value ranges of Climatic Moisture Deficit (CMD), adjusted for rSMR position can be used to reasonably match an expert assigned aSMR x rSMR grid allowing representative CMD values for each BGC to estimate the aSMR for each rSMR/site series across all BGCv13 units.
Klinka et al. (2000) provides tree species frequency ratings by aSMR. We use the driest aSMR with a frequency rating of '1'(very infrequent) to represent where the species is at its drought tolerance limit. To check that the CMD estimated tree species aSMR limits are reasonable we compare them to the tree species frequency ratings from Klinka et al.(2000). 

We do a similar analysis using ForDrat model as a comparison.

We then choose an aSMR level that is considered drought prone. This is set at aSMR <= 2.5. This is based on the assumption that aSMR < 2.5 is drought prone and aSMR > 2.5 is not drought prone under typical climatic conditions. This is a subjective choice and should be reviewed.
To account for topographic effect on local climate we drought conditions on insolation and cold aspect sites. For this we adjust the CMD values for aspect using some approximations derived from Jackson (1966)and elevation. Insolation slopes have CMD * 1.25 and Cold aspects CMD * 0.85. We could use edge cases where aspect and elevation are noted as reforestation considerations to test these predictions (suitable - nonsuitable, aspect restricted sites, elevation restricted sites, etc.)
The adjustment is based on the assumption that the CMD values are too high for sites with a south aspect and too low for sites with a north aspect. This is a subjective choice and should be reviewed.
Some  Does this rule match the environmental suitablility Hold out data and then test against all site series import suitability tables, edaphic tables and find driest site series by BGC and driest rSMR

```{r min suit CMD}
suit <- fread("./inputs/suitability.csv") %>% filter(!newsuit %in% c("4", "5"), !is.na(newsuit))
edatopic <- fread("./inputs/edatopic.csv") %>% 
  mutate(rSMR = as.numeric(substr(Edatopic, 2, 2))) %>% 
  filter(startsWith(Source, "BEC"))

driest.ss <- edatopic %>% 
  group_by(SS_NoSpace) %>% 
  slice_min(order_by = rSMR, n = 1) %>% 
  ungroup() %>% 
  select(SS_NoSpace, rSMR) %>% distinct %>% rename(ss_nospace = SS_NoSpace)

min.suit <- merge(suit, driest.ss, by = "ss_nospace") %>% 
  select(bgc, ss_nospace, spp, newsuit, rSMR) %>% 
  group_by(bgc, spp) %>% 
  slice_min(order_by = rSMR, n = 1) %>% 
  ungroup() %>% distinct %>% mutate(BGC = bgc)
  #select(ss_nospace, spp, newsuit, rSMR) %>% 
#siteseries <- fread("./inputs/site_series.csv")
rsmr_cmd.long <- rsmr_cmd.long %>% filter(PERIOD == "1961_1990", cmd.type == "CMD") %>% 
  mutate(rSMR = as.numeric(substr(rSMR, nchar(rSMR), nchar(rSMR)))) %>% ungroup %>%  data.table

# rsmr_cmd.long <- rsmr_cmd.long %>%
#   rename(bgc = BGC) %>%
#   mutate(rSMR = as.numeric(rSMR))  # Ensure numeric type

min.suit.cmd <- merge(min.suit, rsmr_cmd.long, by = c("BGC", "rSMR")) %>% 
  select(bgc, ss_nospace, spp, newsuit, rSMR, CMD.site, aSMR) %>% 
  group_by(bgc, spp) %>% 
  slice_max(order_by = CMD.site, n = 1) %>% 
  ungroup() %>% distinct %>% filter(newsuit %in% c("2","3"))
```

### Range of CMD from driest rSMR sites for each tree species
```{r min CMD ranges}
#ggplot boxplot of CMD.site by spp
min.suit.cmd %>% 
  ggplot(aes(x = spp, y = CMD.site)) +
  geom_boxplot() +
  theme_minimal() +
  labs(title = "Boxplot of CMD by Species", x = "Species", y = "CMD") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

min.suit.cmd <- min.suit.cmd %>%
  left_join(asmr.codes, by = "aSMR") 
# setDT(min.suit.cmd)[setDT(asmr.codes), aSMR_code := aSMR.code, on = "aSMR"]
# Adds `aSMR_code` column
asmr.codes.used <- min.suit.cmd %>% pull(aSMR.code) 
# Merge with `klinka.min` dataset

#ggplot boxplot of aSMR by spp
```
### Species limits from CMD model compared to Klinka

```{r min aSMR CMD}
klinka.min <- klinka %>% filter(!frequency %in% c("min2", "min3"))
klinka.min <- klinka.min %>%
  left_join(asmr.codes, by = c("min_aSMR" = "aSMR"))

min.suit.cmd %>% 
  ggplot(aes(x = spp, y = aSMR)) +
  geom_boxplot() +
    geom_point(data = klinka.min, aes(x = spp, y = min_aSMR), color = "red", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Boxplot of driest aSMR from CMD by Species ", x = "Species", y = "aSMR",
       subtitle = "red dots are Klinka aSMR minimums") +
  #scale_y_continuous(breaks = 0:9, labels = aSMR.code) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#xx <- min.suit.cmd %>% filter(spp == "Fd") %>% arrange (desc(aSMR))
# min.suit.cmd %>% 
#   ggplot(aes(x = spp, y = aSMR.code)) +  # Change y-axis to aSMR_code
#   geom_boxplot() +
#   geom_point(data = klinka.min, aes(x = spp, y = aSMR.code), color = "red", alpha = 0.7) +  # Use aSMR_code for points
#   theme_minimal() +
#   labs(title = "Boxplot of driest aSMR from CMD by Species", x = "Species", y = "aSMR Code",
#        subtitle = "Red dots are Klinka aSMR minimums") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

# klinka %>% filter(!frequency %in% c("min2", "min3")) %>% 
#   ggplot(aes(x = spp, y = min_aSMR)) +
#   geom_boxplot() +
#   theme_minimal() +
#   labs(title = "Boxplot of Klinka stressed aSMR by Species", x = "Species", y = "aSMR") +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))
```
### Species limits from ForDrat compared to Klinka

```{r fordat min aSMR}

  #select(ss_nospace, spp, newsuit, rSMR) %>% 
#siteseries <- fread("./inputs/site_series.csv")
fordrat.long <- fordrat.long %>%
  mutate(rSMR = as.numeric(substr(rSMR, nchar(rSMR), nchar(rSMR))))

min.suit2 <- merge(min.suit, fordrat.long, by = c("bgc", "rSMR")) %>% 
  select(bgc, ss_nospace, spp, newsuit, rSMR, aSMR.fordrat) %>% 
  group_by(bgc, spp) %>% 
  slice_max(order_by = aSMR.fordrat, n = 1) %>% 
  ungroup() %>% distinct

#ggplot boxplot of CMD.site by spp
min.suit2 %>% 
  ggplot(aes(x = spp, y = aSMR.fordrat)) +
  geom_boxplot() +
      geom_point(data = klinka.min, aes(x = spp, y = min_aSMR), color = "red", alpha = 0.7) +
  theme_minimal() +
  labs(title = "Boxplot of lowest aSMR by Species from ForDrat ", x = "Species", y = "aSMR",
       subtitle = "red dots are Klinka aSMR minimums") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# convert rSMR to CMD
```
## Identify Drought prone Site Series
We choose site series with aSMR <= 2.5 to be considered drought prone.
```{r}
rsmr_cmd.long <- fread("./outputs/modelled_aSMR_by_rSMR_long.csv") %>% 
  mutate(rSMR = as.numeric(substr(rSMR, nchar(rSMR), nchar(rSMR))))

rsmr_cmd.modern <- rsmr_cmd.long %>% filter(PERIOD == "2001_2020")
rsmr_cmd.normal <- rsmr_cmd.long %>% filter(PERIOD == "1961_1990")
rsmr_cmd.2021 <- rsmr_cmd.long %>% filter(PERIOD == "2021")

id.drought.prone = function(rsmr_cmd, period, asmr.limit) {
  # Filter for drought prone sites
  drought.prone <- rsmr_cmd %>% filter(PERIOD == period, aSMR <= asmr.limit) %>% left_join(edatopic, by = c("BGC", "rSMR")) %>% distinct(SS_NoSpace, cmd.type) %>% mutate(drought = "y") %>% ungroup() %>%
    filter(!is.na(SS_NoSpace)) %>% pivot_wider(id_cols = SS_NoSpace, names_from = cmd.type, values_from = drought) %>% arrange(SS_NoSpace) %>% replace_na(list(CMD = "ok", CMD.hot = "ok", CMD.cold = "ok")) %>% filter(!str_detect(SS_NoSpace, "Gg|Ro|Gs|Rt|Gb|00|/x|Bb|Br|Fm")) %>% 
    rename(
      `SiteSeries` = SS_NoSpace,
      `Neutral` = CMD,
      `Insolation` = CMD.hot,
      `CoolAspect` = CMD.cold
    )
  return(drought.prone)
}

drought_normal <- id.drought.prone(rsmr_cmd.long, period = "1961_1990", asmr.limit = 2.5)
drought_modern <- id.drought.prone(rsmr_cmd.long, period = "2001_2020", asmr.limit = 2.5)
drought_2021 <- id.drought.prone(rsmr_cmd.long, period = "2021", asmr.limit = 2.5)
```

`#show figure: set block(breakable: true)`{=typst}

```{r  table report}

gt::gt(drought_modern) %>%
  gt::tab_stubhead(label = "Site Series") %>%
  gt::tab_header(
    title = "Drought Prone Site Series (2001-2020)",
    subtitle = "aSMR <= 2.5"
  ) %>% 
  gt::tab_options(
    column_labels.font.size = 12,
    column_labels.font.weight = "bold",
    table.font.size = 12,
    table.font.weight = "bold"
  ) %>%
  gt::cols_label(
    SiteSeries = "Site Series",
    Neutral = "Neutral",
    Insolation = "Insolation",
    CoolAspect = "Cool Aspect"  
) %>% gt::cols_move(columns = "Insolation", after = "SiteSeries")
```

```{r compare list of drought}
comparison <- drought_modern %>%
  full_join(drought_2021, by = "SiteSeries") %>%
  mutate(flag_difference = ifelse(Neutral.x != Neutral.y, "Mismatch",
                           ifelse(Insolation.x != Insolation.y, "Mismatch",       ifelse(CoolAspect.x != CoolAspect.y, "Mismatch", "Match")))) %>% filter(flag_difference == "Mismatch")
new.modern.drought.ss <- anti_join(drought_modern, drought_normal, by = "SiteSeries")
extreme.drought.ss <- anti_join(drought_2021, drought_modern, by = "SiteSeries")

gt::gt(extreme.drought.ss) %>%
  gt::tab_stubhead(label = "Site Series") %>%
  gt::tab_header(
    title = "Additional Drought Prone Site Series in heat dome year (2021)",
    subtitle = "aSMR <= 2.5"
  ) %>% 
  gt::tab_options(
    column_labels.font.size = 12,
    column_labels.font.weight = "bold",
    table.font.size = 12,
    table.font.weight = "bold"
  ) %>%
  gt::cols_label(
    SiteSeries = "Site Series",
    Neutral = "Neutral",
    Insolation = "Insolation",
    CoolAspect = "Cool Aspect"
)%>% gt::cols_move(columns = "Insolation", after = "SiteSeries")
```


```{r estimate CMD.site limits}
# initial limits
# tree.max.cmd <- read.csv("./inputs/treespecies_CMD_limit.csv")
# CMD <- fread("./outputs/CMD_by_rSMR_matrix.csv")
# rsmr_cmd.long <- CMD %>% pivot_longer(cols = -"BGC", names_to = "excessive.rSMR", values_to = "CMD.site") %>%
#   mutate(CMD.insolation = CMD.site * 1.5)# %>% # adjust for aspect
# # filter to only BGC starting with IDF
# #rsmr_cmd.long <- rsmr_cmd.long %>% filter(grepl("^(CDF|CMA|CWH|MH|IDFww|ICHun|ESSFun1|MSun)", BGC))
# rsmr_cmd.long <- rsmr_cmd.long %>% filter(grepl("^(IDF|PP)", BGC))
# # Loop through all unique species in tree.max.cmd
# tree.max.cmd %>% 
#   distinct(spp) %>% 
#   pull(spp) %>% 
#   walk(~{
#     tree.choose <- .x
#     #tree.choose <- "Sx"
#     tree.choose.cmd <- tree.max.cmd %>% filter(spp == tree.choose) %>% pull(cmd.max)
#     
#     max.rsmr_cmd <- rsmr_cmd.long %>% 
#       filter(CMD.site <= tree.choose.cmd) %>% 
#       group_by(BGC) %>%  
#       slice_max(order_by = CMD.site, n = 1) %>% 
#       ungroup()
#     
#     # Generate table for each species
#     table_output <- gt::gt(max.rsmr_cmd) %>%
#       gt::tab_stubhead(label = "BGC") %>%
#       gt::tab_header(
#         title = paste0("minimum rSMR class for ", tree.choose, " in each BGC"),
#         subtitle = "CMD values adjusted for rSMR position"
#       ) %>% 
#       gt::tab_options(
#         column_labels.font.size = 12,
#         column_labels.font.weight = "bold",
#         table.font.size = 12,
#         table.font.weight = "bold"
#       )
# 
# gt::gtsave(table_output, path = "./outputs/", file = paste0(tree.choose, 'min_rSMR.png'))
#  #   print(table_output)  # Show the table
#   })
# 

```
